Optimal analysis of Discrete-time Time-Invariant Affine Systems

### 2018-10-17 paper.tex
version finale?

 We recall that we have :
%Let us consider the same problem where we replace the linear dynamics by an affine one:
\[
x_0\in \xin,\ \forall\, k\in\nn,\ x_{k+1}=A x_k+b,\
\]
where $A$ is a $d\times d$ matrix and $b$ a vector of $\rd$. We recall that $\rea$ is given by Formula~\eqref{reach} and we consider the optimization problem:
\[
\sup_{x\in\rea} x^\intercal Q x+q^\intercal x
\]
Assumption~\ref{assum2} still holds and it implies that $Id-A$ is invertible. It is well-known that the sequence defined by :
\[
\forall\, k\in\nn,\ y_k=x_k-B,\text{ where } B=(Id-A)^{-1} b
\]
satisfies:
\[
\forall\, k\in\nn,\ y_{k+1}=A y_k. 
\]
Finally, we have :
\[
\forall\, k\in\nn,\ x_k=A^k y_0+B
\]
This latter expression leads to a new formulation of Problem~\eqref{pbopt}: 
\[
\sup_{k\in\nn} \sup_{y_0\in\xin-B} \left(A^k y_0+B\right)^\intercal Q  \left(A^k y_0+B\right)+q^\intercal  \left(A^k y_0+B\right)
\]
or
\[
\sup_{k\in\nn} \sup_{y_0\in\xin-B} y_0^\intercal (A^k)^\intercal Q  A^k y_0+(2B+q)^\intercal A^k +B^\intercal Q B+q^\intercal B
\]
We conclude that we can use the results developed in Subsection~\ref{mainsub} where the matrix $Q$ is unchanged and the vector $q$ to use is now $2B+q$. The polytope of initial conditions also changes since we have to consider now $\xin-B$. We have to adapt Assumption~\ref{assumR} with the new linear part equal to $2B+q$ and the new initial polytope equal to $\xin-B$.

@@

\subsubsection{Homogeneous objective function}
\label{exp-linear}
We consider the discretisation of an harmonic oscillator $\ddot{x}+\dot{x}+x=0$ by an explicit Euler scheme. The discretization step $h$ is set to 0.01. Introducing the position variable, $x$ and the speed variable $v$. We assume that the initial conditions can be taken into the set $[-1,1]^2$. The Euler scheme becomes a linear discrete-time system in dimension two defined as follows: 
\[
\begin{pmatrix}
x_{k+1}\\
v_{k+1}
\end{pmatrix}
=\begin{pmatrix}
1 & h\\
-h & 1-h
\end{pmatrix}
\begin{pmatrix}
x_{k}\\
v_{k}
\end{pmatrix},\ (x_0,v_0)\in [-1,1]^2
\]
We want to prove that :
the trajectories are bounded and compute a bound over the Euclidean norm of the state-variable; for all $k\in\nn$, $x_{k}^2$ and $v_{k}^2$ are less than 1.

\begin{enumerate}
\item Boundedness. The boundedness is already proved since $\rho{A}<1$ and $A^k x$ tends to 0 as $k$ tends to $+\infty$. However, we do not have a bound over the value taken by the state-variable.  
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{6}{|c|}{$\norm{(x_k,v_k)}^2$}\\
\hline
Numbers/Objective functions & $F_0$ & $F_{1}$ & $F_{2}$ & $F_3$ & $F_4$ \\
\hline
$K_1$ & & & & & \\
\hline
 $K_{\rm max}$ & & & & & \\
\hline
\end{tabular}
\end{center}

@@

\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{6}{|c|}{$x_k^2$}\\
\hline
Numbers/Objective functions & $F_0$ & $F_{1}$ & $F_{2}$ & $F_3$ & $F_4$ \\
\hline
$K_1$ & & & & & \\
\hline
 $K_{\rm max}$ & & & & & \\
\hline
\end{tabular}
\end{center}

@@

\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{6}{|c|}{$v_k^2$}\\
\hline
Numbers/Objective functions & $F_0$ & $F_{1}$ & $F_{2}$ & $F_3$ & $F_4$ \\
\hline
$K_1$ & & & & & \\
\hline
 $K_{\rm max}$ & & & & & \\
\hline
\end{tabular}
\end{center}
\end{enumerate}

\subsubsection{Non-Homogeneous objective function}
Let us consider the same linear system as in Subsubsection~\ref{exp-linear}. We are interested in proving that for all $k\in\nn$, $-1\leq x_k-0.5v_k\leq 2$. Following Remark~\ref{remarklinear}, this is the same as proving $(x_k-0.5 v_k)^2-x_k+0.5 v_k\leq 1$. Then, we have to compute the optimal value :
\[
\sup_{x\in\xin} \sup_{k\in\nn} x^\intercal {A^k}^\intercal Q A^k x+q^\intercal A^k  
\]
where $Q=\begin{pmatrix} 1 & -1/2 \\ -1/2 & 1/4\end{pmatrix}$ and $q^\intercal=(-1,1/2)$. 
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{6}{|c|}{$x^\intercal {A^k}^\intercal Q A^k x+q^\intercal A^k x$}\\
\hline
Numbers/Objective functions & $F_0$ & $F_{1}$ & $F_{2}$ & $F_3$ & $F_4$ \\
\hline
$K_1$ & & & & & \\
\hline
 $K_{\rm max}$ & & & & & \\
\hline
\end{tabular}
\end{center}
\subsection{Affine systems}

@@

@@

We want to prove that for all $k\in\nn$, $x_k^2\leq 9$ and $y_k^2\leq 9$.

\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{6}{|c|}{$x^2_k$}\\
\hline
Numbers/Objective functions & $F_0$ & $F_{1}$ & $F_{2}$ & $F_3$ & $F_4$ \\
\hline
$K_1$ & & & & & \\
\hline
 $K_{\rm max}$ & & & & & \\
\hline
\end{tabular}
\end{center}

\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{6}{|c|}{$y_k^2$}\\
\hline
Numbers/Objective functions & $F_0$ & $F_{1}$ & $F_{2}$ & $F_3$ & $F_4$ \\
\hline
$K_1$ & & & & & \\
\hline
 $K_{\rm max}$ & & & & & \\
\hline
\end{tabular}
\end{center}

@@

Taking the same system depicted at Eq, we want to prove that $-2\leq x_k+y_k\leq 4$ is invariant by the dynamics. From Remark~\ref{remarklinear}, to prove the property is equivalent to prove $(x_k+ y_k)^2-x_k- y_k\leq 8$. Then, we have to compute the optimal value :
\[
\sup_{x\in\xin} \sup_{k\in\nn} x^\intercal {A^k}^\intercal Q A^k x+q^\intercal A^k x  
\]
where $Q=\begin{pmatrix} 1 & 1 \\ 1 & 1\end{pmatrix}$ and $q^\intercal=(-1,-1)$.

\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{6}{|c|}{$x^\intercal {A^k}^\intercal Q A^k x+q^\intercal A^k x$}\\
\hline
Numbers/Objective functions & $F_0$ & $F_{1}$ & $F_{2}$ & $F_3$ & $F_4$ \\
\hline
$K_1$ & & & & & \\
\hline
 $K_{\rm max}$ & & & & & \\
\hline
\end{tabular}
\end{center}
\section{Conclusion and Future Works}
\label{conclusion}
In this paper, we introduce a new class of optimization problems. This class solves verification problem coming from computer science. This class can be described as follows : the objective function is convex and quadratic and the decision variable is constrained to belong to the reachable values set of a discrete-time affine system which is not a convex set. The main issue is the infinite horizon problem. To maximize the function we can wait for an arbitrary long time. Thus the paper proposes to compute, in the case of stable affine system, a finite horizon after which the search of an optimal value is useless. The problem becomes a finite horizon problem which can be solved in a finite time.

>>>>

 We recall that we have :
%Let us consider the same problem where we replace the linear dynamics by an affine one:
\[
x_0\in \xin,\ \forall\, k\in\nn,\ x_{k+1}=A x_k+b,\
\]
where $A$ is a $d\times d$ matrix and $b$ a vector of $\rd$. We recall that $\rea$ is given by Formula~\eqref{reach} and we consider the optimization problem:
\[
\sup_{x\in\rea} x^\intercal Q x+q^\intercal x
\]
Assumption~\ref{assum2} still holds and it implies that $Id-A$ is invertible. It is well-known that the sequence defined by :
\[
\forall\, k\in\nn,\ y_k=x_k-B,\text{ where } B=(Id-A)^{-1} b
\]
satisfies:
\[
\forall\, k\in\nn,\ y_{k+1}=A y_k. 
\]
Finally, we have :
\[
\forall\, k\in\nn,\ x_k=A^k y_0+B
\]
This latter expression leads to a new formulation of Problem~\eqref{pbopt}: 
\[
\sup_{k\in\nn} \sup_{y_0\in\xin-B} \left(A^k y_0+B\right)^\intercal Q  \left(A^k y_0+B\right)+q^\intercal  \left(A^k y_0+B\right)
\]
or
\[
\sup_{k\in\nn} \sup_{y_0\in\xin-B} y_0^\intercal (A^k)^\intercal Q  A^k y_0+(2B^\intercal Q+q^\intercal) A^k y_0 +B^\intercal Q B+q^\intercal B
\]
We conclude that we can use the results developed in Subsection~\ref{mainsub} where the matrix $Q$ is unchanged and the vector $q$ to use is now $2B+q$. The polytope of initial conditions also changes since we have to consider now $\xin-B$. We have to adapt Assumption~\ref{assumR} with the new linear part equal to $2QB+q$ and the new initial polytope equal to $\xin-B$.

@@

\subsubsection{Homogeneous objective function}
\label{exp-linear}
We consider the discretisation of an harmonic oscillator $\ddot{x}+\dot{x}+x=0$ by an explicit Euler scheme. The discretization step $h$ is set to 0.01. Introducing the position variable, $x$ and the speed variable $v$. We assume that the initial conditions can be taken into the set $[-1,1]^2$. The Euler scheme becomes a linear discrete-time system in dimension two defined as follows: 
\[
\begin{pmatrix}
x_{k+1}\\
v_{k+1}
\end{pmatrix}
=\begin{pmatrix}
1 & h\\
-h & 1-h
\end{pmatrix}
\begin{pmatrix}
x_{k}\\
v_{k}
\end{pmatrix},\ (x_0,v_0)\in [-1,1]^2
\]
We want to prove that :
the trajectories are bounded and compute a bound over the Euclidean norm of the state-variable; for all $k\in\nn$, $x_{k}^2$ and $v_{k}^2$ are less than 1.
Note that Assumption~\ref{assumR} holds as the hypothesis of Prop~\ref{interiorprop} is satisfied.

\begin{enumerate}
\item Boundedness. The boundedness is already proved since $\rho(A)<1$ and $A^k x$ tends to 0 as $k$ tends to $+\infty$. However, we do not have a bound over the values taken by the state-variable. To bound these values, we compute the maximum of the Euclidean norm of the state-variable :   
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{6}{|c|}{$\norm{(x_k,v_k)}^2$}\\
\hline
Numbers/Objective functions & $F_0$ & $F_{1}$ & $F_{2}$ & $F_3$ & $F_4$ \\
\hline
$K_{\rm max}$ &135 &135 &130 &130 &130 \\
\hline
 $K_{1}$ & 169& 169&169 &169 & 169\\
\hline
\end{tabular}
\end{center}

Actually, the maximum of the square of the Euclidean norm is reached at $k=0$ (for initial conditions) and is equal to $2$. Our computations show that we have to stop at $k=130$. Note that the sequence $\sup_{x\in\xin} \norm{A^k x}_2^2$ is not decreasing i.e. $Id$ is not a solution of the discrete Lyapunov matrix equation.

@@

\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{6}{|c|}{$x_k^2$}\\
\hline
Numbers/Objective functions & $F_0$ & $F_{1}$ & $F_{2}$ & $F_3$ & $F_4$ \\
\hline
$K_{\rm max}$ &230 &230 & 230&230 &230 \\
\hline
 $K_{1}$ &188 &188 &188 &188 &188 \\
\hline
\end{tabular}
\end{center}

Actually, the maximum of the square of the first coordinate is reached at $k=61$ and is equal to $1.6489$. Our computations show that we have to stop at $k=188$. This disproves the property i.e. for all $k\geq 0$, $x_k^2\leq 1$ does not hold.

@@

\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{6}{|c|}{$v_k^2$}\\
\hline
Numbers/Objective functions & $F_0$ & $F_{1}$ & $F_{2}$ & $F_3$ & $F_4$ \\
\hline
$K_{\rm max}$ &233 &233 &221 & 221&227 \\
\hline
 $K_{1}$ &261 &261 &261 &261 &261 \\
\hline
\end{tabular}
\end{center}

Actually, the maximum of the square of the second coordinate is reached at $k=0$ (for initial conditions) and is equal to $1$. Our computations show that we have to stop at $k=221$. Consequently, the property holds. Note that the sequence $\sup_{x\in\xin} (A^k x)_2^2$ is not decreasing.
\end{enumerate}

\subsubsection{Non-Homogeneous objective function}
Let us consider the same linear system as in Subsubsection~\ref{exp-linear}. We are interested in proving that for all $k\in\nn$, $-2\leq x_k-0.5v_k\leq 3$. Following Remark~\ref{remarklinear}, this is the same as proving $(x_k-0.5 v_k)^2-x_k+0.5 v_k\leq 6$. Then, we have to compute the optimal value :
\[
\sup_{k\in\nn} x_k^\intercal Q x_k+q^\intercal x_k
\]
where $Q=\begin{pmatrix} 1 & -1/2 \\ -1/2 & 1/4\end{pmatrix}$ and $q^\intercal=(-1,1/2)$. 
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{6}{|c|}{$x^\intercal {A^k}^\intercal Q A^k x+q^\intercal A^k x$}\\
\hline
Numbers/Objective functions & $F_0$ & $F_{1}$ & $F_{2}$ & $F_3$ & $F_4$ \\
\hline
$K_{\rm max}$ & 285&285 &285 &285 &285 \\
\hline
 $K_{1}$ &276 &276 &279 &279 &262 \\
\hline
\end{tabular}
\end{center}

The table shows that we should stop at the iteration $k=262$ to ensure that the maximum of $(x_k-0.5 v_k)^2-x_k+0.5 v_k$ is reached before the step. Actually the maximum is reached at iteration $k=0$ and is equal to $3.75$ which indicates that the property does hold since we wanted to prove that
$(x_k-0.5 v_k)^2-x_k+0.5 v_k\leq 6$. 
\subsection{Affine systems}

@@

In practice, our Matlab implementation found $Q$ as optimal solution for Problem~\eqref{eq:kunp} when the objective functions are $F_i$ with $i=1,2,3,4$ (all of them except $F_0(P)=\sup_{x\in \xin} x^\intercal P x$. For the case of optimal solutions of Problem~\eqref{eq:klmaxp}, $Q$ is never returned. In the two cases, $K(t,P)=1$ whatever the objective functions.

@@

We want to prove that for all $k\in\nn$, $x_k^2\leq 16$ and $y_k^2\leq 16$.

Note that Assumption~\ref{assumR} holds as the hypothesis of Prop~\ref{interiorprop} is satisfied.

\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{6}{|c|}{$x^2_k$}\\
\hline
Numbers/Objective functions & $F_0$ & $F_{1}$ & $F_{2}$ & $F_3$ & $F_4$ \\
\hline
$K_{\rm max}$ & 6&6 & 6&9&6 \\
\hline
 $K_{1}$ & 97& 97& 94&94 &8 \\
\hline
\end{tabular}
\end{center}
Actually, the maximum of the square of the first coordinate is reached at $k=1$ and is equal to $10.1483$. Our computations show that we have to stop at $k=6$. This proves the property i.e. for all $k\geq 0$, $x_k^2\leq 16$ does hold.

\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{6}{|c|}{$y_k^2$}\\
\hline
Numbers/Objective functions & $F_0$ & $F_{1}$ & $F_{2}$ & $F_3$ & $F_4$ \\
\hline
$K_{\rm max}$ &11 &11 &11 &12 &11 \\
\hline
 $K_{1}$ &264 & 273&263 & 263&19 \\
\hline
\end{tabular}
\end{center}

Actually, the maximum of the square of the first coordinate is reached at $k=4$ and is equal to $21.1427$. Our computations show that we have to stop at $k=11$. This disproves the property i.e. for all $k\geq 0$, $y_k^2\leq 16$ does not hold.

@@

Taking the same system depicted at Eq, we want to prove that $-7\leq 0.5x_k-2y_k\leq 5$ is invariant by the dynamics. From Remark~\ref{remarklinear}, to prove the property is equivalent to prove $(0.5x_k-2 y_k)^2-x_k+4y_k\leq 35$. Then, we have to compute the optimal value :
\[
\sup_{k\in\nn} x_k^\intercal Q x_k+q^\intercal x_k  
\]
where $Q=\begin{pmatrix} 1/4 & -1 \\ -1 & 4\end{pmatrix}$ and $q^\intercal=(-1,4)$.

\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{6}{|c|}{$x_k^\intercal Q x_k+q^\intercal x_k$}\\
\hline
Numbers/Objective functions & $F_0$ & $F_{1}$ & $F_{2}$ & $F_3$ & $F_4$ \\
\hline
$K_{\rm max}$ & 11&11 &11 &12 &11 \\
\hline
 $K_{1}$ & 27 &1100 &1114 &1114 &18 \\
\hline
\end{tabular}
\end{center}

The table shows that we should stop at the iteration $k=11$ to ensure that the maximum of $(0.5x_k-2 y_k)^2-x_k+4y_k$ is reached before the step $k=11$. Actually the maximum is reached at iteration $k=4$ and is equal to $73.295$ which indicates that the property does hold since we wanted to prove that
$(x_k-0.5 v_k)^2-x_k+0.5 v_k\leq 35$.

\section{Conclusion and Future Works}
\label{conclusion}
In this paper, we introduce a new class of optimization problems. This class solves verification problem coming from computer science. This class can be described as follows : the objective function is convex and quadratic and the decision variable is constrained to belong to the reachable values set of a discrete-time affine system which is not a convex set. The main issue is the infinite horizon problem. To maximize the function we can wait for an arbitrary long time. Thus the paper proposes to compute, in the case of stable affine system, a finite horizon after which the search of an optimal value is useless. The problem becomes a finite horizon problem which can be solved in a finite time.
